An essentially algebraic formulation of dependent type theory

### 2015-06-12 weakeningcats.tex
huge revision

\begin{rmk}
Recall that the slice pre-category $\cat{C}/\Gamma/(\Delta,f)$ is isomorphic to the slice pre-category
$\cat{C}/\Delta$, in the essentially algebraic theory of pre-categories. In the following,
we shall usually identify these two categories. Recall also that for every functor
$F:\cat{C}\to\cat{D}$ and every object $\Delta$ of $\cat{C}$, there is a functor
$F/\Delta:\cat{C}/\Delta\to \cat{D}/F(\Delta)$ given by the action on morphisms of $F$. 
\end{rmk}

\begin{rmk}
It will be convenient to denote the objects of the category $\cat{C}$ with capital
greek letters $\Gamma$, $\Delta$, \ldots, and the morphisms by capital latin letters
$A$, $B$, $P$, $Q$, etcetera. When $A$ is a morphism into $\Gamma$, we will denote
its codomain by $\ctxext{\Gamma}{A}$. When $A$ is a morphism into $\Gamma$ and
$P$ is a morphism into $\ctxext{\Gamma}{A}$, it will be convenient to denote the
composition of $P$ with $A$ by $\ctxext{A}{P}$ as well. Note that $\ctxext{A}{P}$
is indeed the domain of $P$ in the slice category $\cat{C}/\Gamma$. Also, objects
of $\cat{C}$ will always have a capital greek letter in their notation, e.g.~as
in $\ctxext{{\Gamma}{A}}{P}$. 
\end{rmk}

\section{Weakening systems}

\begin{defn}
A \emph{pre-weakening structure on a pre-category $\cat{C}$} consists of a functor
$W_A:\cat{C}/\Gamma\to \cat{C}/\ctxext{\Gamma}{A}$ for every $A:\ctxext{\Gamma}{A}\to \Gamma$ in $\cat{C}$ such 
that
\begin{enumerate}
\item $W_{\catid{\Gamma}}=\catid{\cat{C}/\Gamma}$ for every object $\Gamma\in\cat{C}$.
\item $W_{\ctxext{A}{P}}=W_P\circ W_A$ for every $P\in\cat{C}/\ctxext{\Gamma}{A}$ and $A\in\cat{C}/\Gamma$.
\item $W_A$ preserves the final object, i.e.~$W_A(\catid{\Gamma})=\catid{\ctxext{\Gamma}{A}}$.
\end{enumerate}
A \emph{pre-weakening system} is a pre-category together with a pre-weakening structure.
\end{defn}

@@

\begin{lem}
Let $\cat{C}$ be a pre-weakening system. Then for any object $\Gamma$ of $\cat{C}$,
the slice pre-category $\cat{C}/\Gamma$ is also a pre-weakening system, with the weakening
structure given by $W(\cat{C}/\Gamma)_P=W(\cat{C})_P$ for each $P\in\cat{C}/\ctxext{\Gamma}{A}$
and each $A\in\cat{C}/\Gamma$.
\end{lem}

@@

A useful special case of this property is where $B=\catid{\Gamma}$. Thus, if $W$ is
a weakening system, then the diagram
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\cat{C}/\ctxext{\Gamma}{Q}
  \arrow[r,"W_A/Q"]
  &
\cat{C}/\ctxext{{\Gamma}{A}}{W_A(Q)}
  \\
\cat{C}/\Gamma
  \arrow[u,"W_Q"]
  \arrow[r,swap,"W_A"]
  &
\cat{C}/\ctxext{\Gamma}{A}
  \arrow[u,swap,"W_{W_A(Q)}"]
\end{tikzcd}
\end{equation*}
commutes for every $A\in\cat{C}/\Gamma$ and $Q\in\cat{C}/\Gamma$. 
\end{rmk}

@@

\begin{eg}
If the underlying pre-category is a group, the axioms of a weakening system
obtain the following form: There is a group homomorphism $W_x:G\to G$ for
every $x\in G$, in such a way that $W_e=\catid{G}$ and $W_{xy}=W_y\circ W_x$
for every $x,y\in G$, and for any $x,y\in G$ the diagram
\begin{equation*}
\begin{tikzcd}
G \arrow[r,"W_y"]
  &
G 
  \\
G \arrow[u,"W_x"]
  \arrow[r,swap,"W_y"]
  &
G \arrow[u,swap,"W_{W_y(x)}"]
\end{tikzcd}
\end{equation*}
is required to commute. Thus, a weakening system on a group $G$ consists of
a group homomorphism $W:\op G\to\mathrm{Aut}(G)$ such that $W_a\circ W_b=
W_{W_a(b)}\circ W_a$. The latter requirement is satisfied if $b\cdot a=a\cdot W_a(b)$
for every $a,b\in G$, or equivalently if $[b,a]=b^{-1}\cdot W_a(b)$.
\end{eg}

@@

\begin{proof}
Note that $e^{-1}xe=x$ and $(xy)^{-1}z(xy)=y^{-1}(x^{-1}zx)y$ for any $x,y,z\in G$.
Also, the diagram
\begin{equation*}
\begin{tikzcd}
G \arrow[r,"\mathrm{conj}_y"]
  &
G 
  \\
G \arrow[u,"\mathrm{conj}_x"]
  \arrow[r,swap,"\mathrm{conj}_y"]
  &
G \arrow[u,swap,"\mathrm{conj}_{y^{-1}xy}"]
\end{tikzcd}
\end{equation*}
commutes.
\end{proof}

@@

\begin{lem}
Let $G$ be a graph, and let $\cat{C}_G$ be the free pre-category generated by
$G$. A pre-weakening structure on $\cat{C}_G$ is determined by functors
$W_e:\cat{C}_G/\Gamma\to\cat{C}_G/\Delta$ for every edge $e:\Delta\to \Gamma$ in $G$, with the
property that $W_e(\catid{\Gamma})=\catid{\Delta}$.

@@

\begin{eg}
Let $G$ be the graph consisting of one vertex $\ast$ and one edge $e:\ast\to\ast$.
Then $\cat{C}_G/\ast=\N$. Then the only weakening structure on $\cat{C}/G$ is
given by $W_e=\catid{\N}$. The reason is that 
\end{eg}

\begin{lem}
Let $P$ be a pre-order. A pre-weakening structure on $P$ is determined by order
preserving maps $W_{a,b}:{\downarrow}(b)\to{\downarrow}(a)$ satisfying
$W_{a,b}(b)=a$, for every $a\leq b$, where
${\downarrow}(x)$ is the set $\{y\in P\mid y\leq x\}$.

@@

\begin{eg}
In a poset with meets, the functions $W_{a,b}:{\downarrow}(b)\to{\downarrow}(a)$
given by $W_{a,b}(x):= a\land x$ assemble a weakening structure, because for
any $a\leq b$ and $x,c'\leq c\leq b$ we have $a\land(c'\land x)=(a\land c')\land
(a\land x)$.  
\end{eg}

\begin{eg}
Consider the poset $(\N,\geq)$, which has a morphism from $a$ to $b$ if $a\geq b$.
On this poset, we may consider the weakening structure defined by $W_{a,b}(x)=
a+(x-b)$.
\end{eg}

\section{Weakening systems with term structure}
\begin{defn}
A \emph{category with a term structure} is a pre-category $\cat{C}$ with a set $T(A)$ for every
$A\in\cat{C}/\Gamma$ for every $\Gamma\in\cat{C}$. A \emph{functor with term structure} from $\cat{C}$ to $\cat{D}$
is a functor $F:\cat{C}\to\cat{D}$ with a function $T(A)\to T(F(A))$ for
every $A\in\cat{C}/\Gamma$, for every $\Gamma\in\cat{C}$.
\end{defn}

\begin{defn}
Suppose $\cat{C}$ is a pre-category with term structure, and $\Gamma$ is an object of $\cat{C}$. 
Then $\cat{C}/\Gamma$ is a pre-category with term structure, where $T_{\cat{C}/\Gamma}(A)=T_\cat{C}(A)$.
\end{defn}

@@

Let $G$ be a group, with the weakening structure given by conjugation. A term
structure on $G$ is given by a set $X$ with a right $G$-action $\mu:\op G\times X\to X$.
We take $T(a):= X$ for every $a\in G$.
\end{eg}

\section{Projection systems}
\begin{defn}
A \emph{pre-projection system} is a weakening system with term structure with an element
$\tfid{A}\in T(W_A(A))$ for every $A\in\cat{C}/\Gamma$ and $\Gamma\in\cat{C}$. 
\end{defn}

\begin{defn}
A \emph{pre-projection homomorphism} $F:\cat{C}\to\cat{D}$ is a weakening homomorphism for which
\begin{equation*}
F(\tfid{A})=\tfid{F(A)}
\end{equation*}
for every $A\in\cat{C}/\Gamma$ and $\Gamma\in\cat{C}$.
\end{defn}

@@

@@

\begin{eg}
Consider a group $G$ with a right $G$-set $X$. A projection structure on $(G,X)$
consists of an element $\tfid{a}\in X$ for every $a\in G$, so that $\tfid{a}\cdot b=
\tfid{b^{-1}ab}$ for every $a,b\in G$.

%For instance, this is the case if we take $X$ to be a pointed set with $\ast\in X$,
%and take $\tfid{a}:=\ast$ for all $a\in G$.
\end{eg}

\section{Substitution systems}

\begin{defn}
A \emph{pre-substitution structure} on a pre-category with term structure $\cat{C}$ consists
of a functor with term structure $S_x:\cat{C}/\ctxext{\Gamma}{A}\to\cat{C}/\Gamma$ for every $x\in T(A)$ and
$A\in\cat{C}/\Gamma$, such that $S_x(\catid{\ctxext{\Gamma}{A}})=\catid{\Gamma}$.

@@

\begin{lem}
For any object $\Delta$ of a pre-substitution system $\cat{C}$, the slice pre-category
$\cat{C}/\Delta$ is a pre-substitution system.
\end{lem}

@@

\begin{cor}
For any object $\Delta$ of a substitution system, the slice pre-category $\cat{C}/\Delta$
is a substitution system.
\end{cor}

@@

\begin{eg}
Let $G$ be a group and let $H$ be a set, forming a category with term 
structure of which $T(a)=H$ for every $a\in G$. A substitution
system on $(G,H)$ consists of a function $\sigma_G : H\times G\to G$ and
a function $\sigma_H:H\times H\to H$, for which the diagrams
\begin{equation*}
\begin{tikzcd}[column sep=large]
G \arrow[r,"\sigma_G(h)"]
  \arrow[d,swap,"\sigma_G(k)"]
  &
G \arrow[d,"{\sigma_G(\sigma_H(h,k))}"]
  \\
G \arrow[r,swap,"\sigma_G(h)"] 
& G
\end{tikzcd}
\end{equation*}
and
\begin{equation*}
\begin{tikzcd}[column sep=large]
H \arrow[r,"\sigma_H(h)"]
  \arrow[d,swap,"\sigma_H(k)"]
  &
H \arrow[d,"{\sigma_H(\sigma_H(h,k))}"]
  \\
H \arrow[r,swap,"\sigma_H(h)"]
  &
H
\end{tikzcd}
\end{equation*}
commute for every $h,k\in H$. 
\end{eg}

\section{E-systems}
\begin{defn}
\label{defn:esystems}
An \emph{E-system} is a pre-category $\cat{C}$ with term structure with
\begin{itemize}
\item a weakening structure $W$,
\item the structure $\tfid{}$ of a projection system,
\item a substitution structure $S$,
\end{itemize}
such that
\begin{enumerate}
\item each $W_A$ is a substitution homomorphism, i.e.~the diagram
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\cat{C}/\ctxext{{\Gamma}{B}}{Q}
  \arrow[r,"W_A/\ctxext{B}{Q}"]
  \arrow[d,swap,"S_y"]
  &
\cat{C}/\ctxext{{{\Gamma}{A}}{W_A(B)}}{W_A(Q)}
  \arrow[d,"S_{W_A(y)}"]
  \\
\cat{C}/\ctxext{\Gamma}{B}
  \arrow[r,swap,"W_A/B"]
  &
\cat{C}/\ctxext{{\Gamma}{A}}{W_A(B)}
\end{tikzcd}
\end{equation*}
commutes for every $Q\in\cat{C}/\ctxext{\Gamma}{B}$, $B\in\cat{C}/\Gamma$ and each 
$y\in T(Q)$; 
\item each $S_x$ is a projection homomorphism, i.e.~the diagram
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\cat{C}/\ctxext{{\Gamma}{A}}{P}
  \arrow[r,"S_x/P"]
  \arrow[d,swap,"W_Q"]
  &
\cat{C}/\ctxext{\Gamma}{S_x(P)}
  \arrow[d,"W_{S_x(Q)}"]
  \\
\cat{C}/\ctxext{{{\Gamma}{A}}{P}}{Q}
  \arrow[r,swap,"S_x/\ctxext{P}{Q}"]
  &
\cat{C}/\ctxext{{\Gamma}{S_x(P)}}{S_x(Q)}
\end{tikzcd}
\end{equation*}
commutes for every $Q\in\cat{C}/\ctxext{{\Gamma}{A}}{P}$, and we have
\begin{equation*}
S_x(\tfid{Q})=\tfid{S_x(Q)}
\end{equation*}
for every $Q\in\cat{C}/\ctxext{{\Gamma}{A}}{P}$;
\item $S_x(\tfid{A})=x$ for any $x\in T(A)$ and $A\in\cat{C}/\Gamma$;
\item \label{tTf:StWf_id} $S_x\circ W_A=\catid{\cat{C}/\Gamma}$ for any $x\in T(A)$ and $A\in\cat{C}/\Gamma$;
\item \label{f:SidfWff_id} $S_{\tfid{A}}\circ W_A/A=\catid{\cat{C}/\ctxext{\Gamma}{A}}$ for any $A\in\cat{C}/\Gamma$.
\end{enumerate}
An \emph{E-homomorphism} $F:\cat{C}\to\cat{D}$ is a functor which is both a projection
homomorphism and a substitution homomorphism.
\end{defn}

\begin{cor}
For any object $\Delta$ of an E-system $\cat{C}$, the slice pre-category $\cat{C}/\Delta$
is an E-system.
\end{cor}

\begin{rmk}[Needs more precision]
Consider the category $E$ generated by the graph consisting of the objects of 
the underlying category of an $E$-system $\mathcal{E}$ as edges, and the morphisms
$S_t:X\to Y$ and $W_A:Y\to X$ as edges, subject to the axioms of an $E$-systems.

Any morphism of $E$ can be rewritten in a unique way
as a composition $\tilde S\circ \tilde W$, where $\tilde S$ is a morphism in
substitution-normal form, $\tilde W$ is in weakening normal form, every term
expression is in normal form, and where
each occurence of $S_x\circ W_A$ and $S_{\tfid{A}}\circ W_A/A$ is eliminated.
\end{rmk}

\begin{defn}
An E-system of which the underlying category has a terminal object $\emptyc$, is said to
be \emph{E-system with empty context}. An E-homomorphism is said to preserve the
empty context when it preserves the terminal object $\emptyc$ on the nose.
\end{defn}

\begin{rmk}
When there is a terminal object, we get a weakening functor $W_\Gamma:\cat{C}\to
\cat{C}/\Gamma$ for every object $\Gamma$ of $\cat{C}$; we get a set $T(\Gamma)$;
and for $i\in T(\Gamma)$ we get substitutions $S_i:\cat{C}/\Gamma\to\cat{C}$. 
\end{rmk}

\section{Properties of E-systems}

\subsection{The category of internal functions}

\begin{defn}
For every $A,B\in\cat{C}/\Gamma$ we define the set 
\begin{equation*}
\thom{A}{B}:=T(W_A(B)).
\end{equation*} 
\end{defn}

\begin{rmk}
Note that $\thom{\catid{\Gamma}}{A}=T(A)$ for any $A\in\cat{C}/\Gamma$, 
because we have $W_{\catid{\ctxext{\Gamma}{A}}}=\catid{\cat{C}/\ctxext{\Gamma}{A}}$.

Note also that $\thom{\ctxext{A}{P}}{B}=\thom{P}{W_A(B)}$ for any $P\in\cat{C}/\ctxext{\Gamma}{A}$
and $B\in\cat{C}/\Gamma$, because $W_{\ctxext{A}{P}}=
W_P\circ W_A$.
\end{rmk}

\begin{defn}
Let $A,B\in\cat{C}/\Gamma$.
For any $f\in\thom{A}{B}$ we define the pre-composition E-homomorphism
\begin{equation*}
f_\ast := S_f\circ W_A/B : \cat{C}/\ctxext{\Gamma}{B}\to\cat{C}/\ctxext{\Gamma}{A}.
\end{equation*}
Furthermore, for $g\in\thom{B}{C}$ we may consider $W_B(C)\in\cat{C}/\ctxext{\Gamma}{B}$ as a morphism from $W_B(C)$ to $\catid{\ctxext{\Gamma}{B}}$. Therefore, $f_\ast$ acts on the terms of
$W_B(C)$, so it makes sense to define $g\circ f:=f_\ast(g)$. 
\end{defn}

Note that $g\circ f\in T(S_f(W_A/B(W_B(C))))$, whereas we would like that
$g\circ f\in\thom{A}{C}$. The following lemma confirms this:

\begin{lem}
Suppose $A$ and $B$ are morphisms of the underlying pre-category of $\mathcal{E}$
with the same codomain, and let $f\in\thom{A}{B}$. Then we have 
\begin{equation*}
S_f\circ (W_A/B)\circ W_B=W_A.
\end{equation*}
\end{lem}

\begin{proof}
Since weakening is a weakening homomorphism, we have
\begin{equation*}
S_f\circ (W_A/B)\circ W_B=S_f\circ W_{W_A(B)}\circ W_A
\end{equation*}
By \autoref{tTf:StWf_id} of \autoref{defn:esystems} it we get that
\begin{equation*}
S_f\circ W_{W_A(B)}\circ W_A=W_A.\qedhere
\end{equation*}
\end{proof}

\begin{rmk}
Note that \autoref{f:SidfWff_id} of \autoref{defn:esystems} asserts precisely
that ${(\tfid{A})}_\ast=\catid{\cat{C}/\Delta}$ for any $A:\Delta\to \Gamma$ in $\cat{C}$. In
particular, it follows that $g\circ\tfid{A}=g$ for any $g\in\thom{A}{B}$
\end{rmk}

\begin{lem}
For any $f\in\thom{A}{B}$ we have $\tfid{B}\circ f=f$.
\end{lem}

\begin{proof}
We have
\begin{equation*}
\tfid{B}\circ f=S_f(W_A(\tfid{B}))=S_f(\tfid{W_A(B)})=f.\qedhere
\end{equation*}
\end{proof}

>>>>

\section{The definition of E-systems}\label{sec:esys_defn}
\subsection{Weakening systems}

\begin{defn}
A \emph{pre-weakening structure on a pre-category $\cat{C}$} consists of a functor
$W_A:\cat{C}/\Gamma\to \cat{C}/\ctxext{\Gamma}{A}$ for every $A:\ctxext{\Gamma}{A}\to \Gamma$ in $\cat{C}$ such 
that
\begin{enumerate}
\item $W_{\catid{\Gamma}}\jdeq \catid{\cat{C}/\Gamma}$ for every object $\Gamma\in\cat{C}$.
\item $W_{\ctxext{A}{P}}\jdeq W_P\circ W_A$ for every $P\in\cat{C}/\ctxext{\Gamma}{A}$ and $A\in\cat{C}/\Gamma$.
\item $W_A$ preserves the final object, i.e.~$W_A(\catid{\Gamma})\jdeq \catid{\ctxext{\Gamma}{A}}$.
\end{enumerate}
A \emph{pre-weakening system} is a pre-category together with a pre-weakening structure.
\end{defn}

@@

\begin{lem}
Let $\cat{C}$ be a pre-weakening system. Then for any object $\Gamma$ of $\cat{C}$,
the slice pre-category $\cat{C}/\Gamma$ is also a pre-weakening system, with the weakening
structure given by $W(\cat{C}/\Gamma)_P\jdeq W(\cat{C})_P$ for each $P\in\cat{C}/\ctxext{\Gamma}{A}$
and each $A\in\cat{C}/\Gamma$.
\end{lem}

@@

A useful special case of this property is where $B\jdeq \catid{\Gamma}$. Thus, if $W$ is
a weakening system, then the diagram
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\cat{C}/\ctxext{\Gamma}{C}
  \arrow[r,"W_A/C"]
  &
\cat{C}/\ctxext{{\Gamma}{A}}{W_A(C)}
  \\
\cat{C}/\Gamma
  \arrow[u,"W_C"]
  \arrow[r,swap,"W_A"]
  &
\cat{C}/\ctxext{\Gamma}{A}
  \arrow[u,swap,"W_{W_A(C)}"]
\end{tikzcd}
\end{equation*}
commutes for every $A,C\in\cat{C}/\Gamma$. In particular, we see
that $W_A(W_C(D))\jdeq W_{W_A(C)}(W_A(D))$ for any $D\in\cat{C}/\Gamma$,
i.e.~that weakening is a self-distributive operation. 
\end{rmk}

@@

\begin{eg}
If the underlying pre-category is a group, the axioms of a weakening system
obtain the following form: There is a group homomorphism $W_x:G\to G$ for
every $x\in G$, in such a way that $W_e\jdeq \catid{G}$ and $W_{xy}\jdeq W_y\circ W_x$
for every $x,y\in G$, and for any $x,y\in G$ the diagram
\begin{equation*}
\begin{tikzcd}
G \arrow[r,"W_y"]
  &
G 
  \\
G \arrow[u,"W_x"]
  \arrow[r,swap,"W_y"]
  &
G \arrow[u,swap,"W_{W_y(x)}"]
\end{tikzcd}
\end{equation*}
is required to commute. Thus, a weakening system on a group $G$ consists of
a group homomorphism $W:\op G\to\mathrm{Aut}(G)$ such that $W_a\circ W_b\jdeq 
W_{W_a(b)}\circ W_a$. The latter requirement is satisfied if $b\cdot a\jdeq a\cdot W_a(b)$
for every $a,b\in G$, or equivalently if $[b,a]\jdeq b^{-1}\cdot W_a(b)$.
\end{eg}

@@

\begin{proof}
Note that $e^{-1}xe\jdeq x$ and $(xy)^{-1}z(xy)\jdeq y^{-1}(x^{-1}zx)y$ for any $x,y,z\in G$.
Also, the diagram
\begin{equation*}
\begin{tikzcd}
G \arrow[r,"\mathrm{conj}_y"]
  &
G 
  \\
G \arrow[u,"\mathrm{conj}_x"]
  \arrow[r,swap,"\mathrm{conj}_y"]
  &
G \arrow[u,swap,"\mathrm{conj}_{y^{-1}xy}"]
\end{tikzcd}
\end{equation*}
commutes.
\end{proof}

@@

\begin{comment}
\begin{lem}
Let $G$ be a graph, and let $\cat{C}_G$ be the free pre-category generated by
$G$. A pre-weakening structure on $\cat{C}_G$ is determined by functors
$W_e:\cat{C}_G/\Gamma\to\cat{C}_G/\Delta$ for every edge $e:\Delta\to \Gamma$ in $G$, with the
property that $W_e(\catid{\Gamma})\jdeq \catid{\Delta}$.

@@

\begin{eg}
Let $G$ be the graph consisting of one vertex $\ast$ and one edge $e:\ast\to\ast$.
Then $\cat{C}_G/\ast\jdeq \N$. Then the only weakening structure on $\cat{C}/G$ is
given by $W_e\jdeq \catid{\N}$. The reason is that 
\end{eg}
\end{comment}

\begin{lem}
Let $P$ be a pre-order. A pre-weakening structure on $P$ is determined by order
preserving maps $W_{a,b}:{\downarrow}(b)\to{\downarrow}(a)$ satisfying
$W_{a,b}(b)\jdeq a$, for every $a\leq b$, where
${\downarrow}(x)$ is the set $\{y\in P\mid y\leq x\}$.

@@

\begin{eg}
In a poset with meets, the functions $W_{a,b}:{\downarrow}(b)\to{\downarrow}(a)$
given by $W_{a,b}(x)\defeq a\land x$ assemble a weakening structure, because for
any $a\leq b$ and $x,c'\leq c\leq b$ we have $a\land(c'\land x)\jdeq (a\land c')\land
(a\land x)$.  
\end{eg}

\begin{eg}
Consider the poset $(\N,\geq)$, which has a morphism from $a$ to $b$ if $a\geq b$.
On this poset, we may consider the weakening structure defined by $W_{a,b}(x)\jdeq
a+(x-b)$.
\end{eg}

\subsection{Weakening systems with term structure}
\begin{defn}
A \emph{category with a term structure} is a pre-category $\cat{C}$ with a set $T(A)$ for every
$A\in\cat{C}/\Gamma$ for every $\Gamma\in\cat{C}$. A \emph{functor with term structure} from $\cat{C}$ to $\cat{D}$
is a functor $F:\cat{C}\to\cat{D}$ with a function $T(A)\to T(F(A))$ for
every $A\in\cat{C}/\Gamma$, for every $\Gamma\in\cat{C}$.
\end{defn}

\begin{defn}
Suppose $\cat{C}$ is a pre-category with term structure, and $\Gamma$ is an object of $\cat{C}$. 
Then $\cat{C}/\Gamma$ is a pre-category with term structure, where $T_{\cat{C}/\Gamma}(A)\jdeq T_\cat{C}(A)$.
\end{defn}

@@

Let $G$ be a group, with the weakening structure given by conjugation. A term
structure on $G$ is given by a set $X$ with a right $G$-action $\mu:\op G\times X\to X$.
We take $T(a)\defeq X$ for every $a\in G$.
\end{eg}

\subsection{Projection systems}
\begin{defn}
A \emph{pre-projection system} is a pre-weakening system with term structure with an element
$\tfid{A}\in T(W_A(A))$ for every $A\in\cat{C}/\Gamma$ and $\Gamma\in\cat{C}$. 
\end{defn}

\begin{defn}
A \emph{pre-projection homomorphism} $F:\cat{C}\to\cat{D}$ is a pre-weakening homomorphism for which
\begin{equation*}
F(\tfid{A})\jdeq \tfid{F(A)}
\end{equation*}
for every $A\in\cat{C}/\Gamma$ and $\Gamma\in\cat{C}$.
\end{defn}

@@

\begin{proof}
Straightforward.
\end{proof}

@@

%\begin{eg}
%Consider a group $G$ with a right $G$-set $X$. A projection structure on $(G,X)$
%consists of an element $\tfid{a}\in X$ for every $a\in G$, so that $\tfid{a}\cdot b=
%\tfid{b^{-1}ab}$ for every $a,b\in G$. 
%For instance, this is the case if we take $X$ to be a pointed set with $\ast\in X$,
%and take $\tfid{a}:=\ast$ for all $a\in G$.
%\end{eg}

\subsection{Substitution systems}

\begin{defn}
A \emph{pre-substitution structure} on a pre-category with term structure $\cat{C}$ consists
of a functor with term structure $S_x:\cat{C}/\ctxext{\Gamma}{A}\to\cat{C}/\Gamma$ for every $x\in T(A)$ and
$A\in\cat{C}/\Gamma$, such that $S_x(\catid{\ctxext{\Gamma}{A}})\jdeq\catid{\Gamma}$.

@@

\begin{lem}
For any object $\Gamma$ of a pre-substitution system $\cat{C}$, the slice pre-category
$\cat{C}/\Gamma	$ is a pre-substitution system.
\end{lem}

@@

\begin{cor}
For any object $\Gamma$ of a substitution system, the slice pre-category $\cat{C}/\Gamma$
is a substitution system.
\end{cor}

@@

\begin{eg}
Let $G$ be a group and let $H$ be a set, forming a category with term 
structure of which $T(a)\jdeq H$ for every $a\in G$. A substitution
system on $(G,H)$ consists of a function $\sigma_G : H\times G\to G$ and
a function $\sigma_H:H\times H\to H$, for which the diagrams
\begin{equation*}
\begin{tikzcd}[column sep=large]
G \arrow[r,"\sigma_G(h)"]
  \arrow[d,swap,"\sigma_G(k)"]
  &
G \arrow[d,"{\sigma_G(\sigma_H(h,k))}"]
  \\
G \arrow[r,swap,"\sigma_G(h)"] 
& G
\end{tikzcd}
\end{equation*}
and
\begin{equation*}
\begin{tikzcd}[column sep=large]
H \arrow[r,"\sigma_H(h)"]
  \arrow[d,swap,"\sigma_H(k)"]
  &
H \arrow[d,"{\sigma_H(\sigma_H(h,k))}"]
  \\
H \arrow[r,swap,"\sigma_H(h)"]
  &
H
\end{tikzcd}
\end{equation*}
commute for every $h,k\in H$. 
\end{eg}

\subsection{E-systems}

Although it is now possible to state the definition of an E-system without
any difficulty, it shall be convenient later to also have a category of
pre-E-systems at our disposal. Pre-E-systems shall by default be non-unital,
i.e.~they do not come with a pre-projection system, while E-systems shall
always be assumed to be unital. In the present context, we shall not need
unital pre-E-systems or non-unital E-systems.

\begin{defn}
A \emph{pre-E-system} is a pre-category $\cat{C}$ with term structure
with a pre-weakening structure $W$, a pre-substitution structure $S$ and a terminal
object. A \emph{unital pre-E-system}, or simply \emph{pre-E-system}, is a non-unital pre-E-system
which also has the structure of a pre-projection system. 
\end{defn}

\begin{defn}
A \emph{pre-E-homomorphism} from $\mathbb{E}\to\mathbb{D}$ is a functor
$H:\mathbb{C}_\mathbb{E}\to\mathbb{C}_\mathbb{D}$ between the underlying categories
with term structure, which is a pre-weakening homomorphism, a pre-substitution
homomorphism, and which preserves the terminal object.

A \emph{unital} pre-E-homomorphism is a non-unital pre-E-homomorphism which is 
also a pre-projection homomorphism. 
\end{defn}

\begin{defn}
\label{defn:esystems}
An \emph{E-system} is a unital pre-E-system in which 
\begin{enumerate}
\item each $W_A$ is a unital pre-E-homomorphism,
\item each $S_x$ is a unital pre-E-homomorphism,
\item \label{tTf:StWf_id} weakened families are constant, i.e.~$S_x\circ W_A\jdeq \catid{\cat{C}/\Gamma}$ for any $x\in T(A)$ and $A\in\cat{C}/\Gamma$,
\item $S_x(\tfid{A})\jdeq x$ for any $x\in T(A)$ and $A\in\cat{C}/\Gamma$, and
\item \label{f:SidfWff_id} $S_{\tfid{A}}\circ W_A/A\jdeq \catid{\cat{C}/\ctxext{\Gamma}{A}}$ for any $A\in\cat{C}/\Gamma$.
\end{enumerate}
An \emph{E-homomorphism} $H:\mathbb{E}\to\mathbb{D}$ is a functor from
$\cat{C}_{\mathbb{E}}$ to $\cat{C}_{\mathbb{D}}$ which is both a projection
homomorphism and a substitution homomorphism.
\end{defn}

\begin{rmk}
The condition that each $W_A$ is a substitution homomorphism asserts that
the diagram
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\cat{C}/\ctxext{{\Gamma}{B}}{Q}
  \arrow[r,"W_A/\ctxext{B}{Q}"]
  \arrow[d,swap,"S_y"]
  &
\cat{C}/\ctxext{{{\Gamma}{A}}{W_A(B)}}{W_A(Q)}
  \arrow[d,"S_{W_A(y)}"]
  \\
\cat{C}/\ctxext{\Gamma}{B}
  \arrow[r,swap,"W_A/B"]
  &
\cat{C}/\ctxext{{\Gamma}{A}}{W_A(B)}
\end{tikzcd}
\end{equation*}
of functors with term structure commutes for every 
$Q\in\cat{C}/\ctxext{\Gamma}{B}$, $B\in\cat{C}/\Gamma$ and each $y\in T(Q)$.

Likewise, the condition that each $S_x$ is a weakening homomorphism
asserts that the diagram
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\cat{C}/\ctxext{{\Gamma}{A}}{P}
  \arrow[r,"S_x/P"]
  \arrow[d,swap,"W_Q"]
  &
\cat{C}/\ctxext{\Gamma}{S_x(P)}
  \arrow[d,"W_{S_x(Q)}"]
  \\
\cat{C}/\ctxext{{{\Gamma}{A}}{P}}{Q}
  \arrow[r,swap,"S_x/\ctxext{P}{Q}"]
  &
\cat{C}/\ctxext{{\Gamma}{S_x(P)}}{S_x(Q)}
\end{tikzcd}
\end{equation*}
of functors with term structure commutes for every 
$Q\in\cat{C}/\ctxext{{\Gamma}{A}}{P}$.
\end{rmk}

\begin{cor}
For any object $\Gamma$ of an E-system $\cat{C}$, the slice pre-category $\cat{C}/\Gamma$
is an E-system.
\end{cor}

\begin{comment}
\begin{rmk}[Needs more precision]
Consider the category $E$ generated by the graph consisting of the objects of 
the underlying category of an $E$-system $\mathcal{E}$ as edges, and the morphisms
$S_t:X\to Y$ and $W_A:Y\to X$ as edges, subject to the axioms of an $E$-systems.

Any morphism of $E$ can be rewritten in a unique way
as a composition $\tilde S\circ \tilde W$, where $\tilde S$ is a morphism in
substitution-normal form, $\tilde W$ is in weakening normal form, every term
expression is in normal form, and where
each occurence of $S_x\circ W_A$ and $S_{\tfid{A}}\circ W_A/A$ is eliminated.
\end{rmk}
\end{comment}

\section{Properties of E-systems}\label{sec:esys_props}

The theory of E-systems could be seen as a sort of category theory with only
projections and their sections. We elaborate on this point of view by defining
the category of internal homomorphisms, which Lawvere might have called the
category of terms. 
In the category of internal morphisms over a context $\Gamma$,
a morphism $f:A\to B$ is a term of the constant family $W_A(B)$ over 
$\ctxext{\Gamma}{A}$. Rather than directly defining composition, we will define
for every $f:A\to B$ in context $\Gamma$, a pre-composition operation
$f_\ast$, which is going to be an E-homomorphism from 
$\cat{C}/\ctxext{\Gamma}{B}$ to $\cat{C}/\ctxext{\Gamma}{A}$. This allows us
to consider diagrams of the following form:
\begin{equation*}
\begin{tikzcd}
P \arrow[d,->>] \arrow[r,"g"] & Q \arrow[d,->>] \\
A \arrow[r,swap,"f"] & B
\end{tikzcd}
\end{equation*}
where we consider the morphism $g$ over $f$ to be an internal morphism from 
$P$ to $f_\ast(Q)$ in context $\ctxext{\Gamma}{A}$, capturing the
essence of being a morphism from the family $P$ over $A$ to the family
$Q$ over $B$, reindexed by $f$. 
Whenever we draw such diagrams, a double headed arrow will always
refer to a family, and an ordinary arrow will refer to an internal morphism,
possibly over another internal morphism

A morphism $g$ over $f$ can be vertically composed with $f$ so
that we obtain a morphism $\jvcomp{P}{f}{g}:\ctxext{A}{P}\to\ctxext{B}{Q}$ in 
context $\Gamma$, and it can be horizontally composed with a morphism $k:Q\to R$
over $h:B\to C$ so that we obtain a morphism $\jfcomp{A}{f}{P}{g}{k}:P\to R$ 
over $\jcomp{A}{f}{h}:A\to C$. We will show interchange laws between these
sorts of composition. The interchange laws have as a consequence that if we
assert the commutativity of a large composed diagram.

We then demonstrate how to define a category internal to an E-system. The main
point of this lies mainly in demonstrating how the operations of E-systems can
be used to define a category, because it is no surprise that a category can be
defined in type theory. Nevertheless, it is noteworthy that we need no further 
conditions on E-systems.
This is unlike the situation in category theory, where we need
to assume that the ambient category has finite limits in order to say what an
internal category is. In fact, it is possible to define an E-system internal
to an E-system, just using the algebraic structure of an E-system, and actually
defining an internal category was the hardest part of that definition.

Before we begin this section, we introduce some handy shorthands.
